---
title: "General exploration of gene content in Casei group genomes"
author: "Stijn Wittouck"
date: "15th of June 2017"
output: 
  html_document:
    depth: 2
    number_sections: yes
    theme: united
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r}
library(tidyverse)
library(stringr)
library(vegan)
source(file = "function_make_pcoaTable.R")
```

# Loading the data

We start by reading in the necessary tables. 

```{r}
load(file = "intermediate/V_genome_orthogroup.Robject")
load(file = "intermediate/T_genome.Robject")
load(file = "intermediate/T_orthogroup.Robject")
load(file = "intermediate/T_eggnogFam_functCat.Robject")
T_functCat <- read_tsv(file = "input/functCat_descriptions.tsv")
```

# Summary statistics of casei group and clades

We want the following information for each clade: number of genomes, number of core orthogroups, number of accessory orthogroups, average number of orthogroups per genome, average number of genes per genome. 

```{r}

calculateCladeStatistics <- function(V_genome_orthogroup) {
  
  part1 <- V_genome_orthogroup %>%
    select(genome) %>%
    distinct() %>%
    summarise(nGenomes = n())
  nGenomesTot <- part1[[1]][1]
  
  part2 <- V_genome_orthogroup %>%
    group_by(orthogroup) %>%
    summarise(ngenomes = sum(ngenes != 0)) %>%
    summarise(nCoreOrthogroups = sum(ngenomes >= (0.96 * nGenomesTot)), 
              nAccOrthogroups = sum(ngenomes < (0.96 * nGenomesTot)))
  
  part3 <- V_genome_orthogroup %>%
    group_by(genome) %>%
    summarise(nOrthogroups = sum(ngenes != 0), nGenes = sum(ngenes)) %>%
    summarise(avnOrthogroups = mean(nOrthogroups), avnGenes = mean(nGenes),
              sdnOrthogroups = sd(nOrthogroups), sdnGenes = sd(nGenes))
  
  return(cbind(part1, part2, part3))
  
}

left_join(V_genome_orthogroup, T_genome) %>%
  list(
    ., 
    filter(., clade == "cladeA"),
    filter(., clade == "cladeB"),
    filter(., clade == "cladeC")
  ) %>%
  lapply(FUN = calculateCladeStatistics) %>%
  bind_rows() %>%
  mutate(clade = c("casei group", "clade A", "clade B", "clade C")) %>%
  column_to_rownames(var = "clade")

```

# Visualization of gene content comparisons 

## PCoA for all orthogroups

Let's make a PCoA plot of the overall pairwise distances between genomes in terms of gene content (Bray-Curtis). The colors match the ones used in the tree figures. 

```{r}
# make table with pcoa coordinates for all genomes 
# also add genome info (clade, species)
T_pcoa1 <- make_pcoaTable(V_genome_orthogroup) %>%
  left_join(T_genome)

# same clade colors as in trees
cladeColors <- c("cladeA" = "#e41a1c", "cladeB" = "#377eb8", "cladeC" = "#4daf4a")

# plot and save
ggplot(T_pcoa1, aes(x = V1, y = V2, col = clade, shape = species)) +
  geom_point() +
  scale_color_manual(values = cladeColors) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
ggsave(filename = "results/pcoa_overall.png", height = 20, width = 30, units = "cm")
```

## PCoA per functional category
Next, we can do the same, but per functional category. Some important things to keep in mind:

* It is important to realize that several orthogroups are present in multiple functional categories!! So these are represented in multiple sub-plots. 
* We will remove functional categories with less than 20 orthogroups from the figure because there is too little information to "resolve" all points (genomes) in the pcoa plot. 
* The distances between genomes are percentages, not of a number of orthogroups but of a number of genes!! It would of course be possible to change this to orthogroup presence/absence instead of considering the numbers of genes. This would yield a slightly different graph. 
* The large majority of the orthogroups has no known function! 

```{r}
# make table with genome, orthogroup, ngenes and funct cat
# remark: mutate step is to make sure that orthogroups without
# eggnogFam hit get functional category "S" (function unknown)
T_pcoa2 <- left_join(V_genome_orthogroup, T_orthogroup) %>%
  left_join(T_eggnogFam_functCat) %>%
  select(genome, orthogroup, ngenes, functCat) %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat))

# make table with number of orthogroups per functional category
T_nOrthogroupsPerFunctCat <- T_pcoa2 %>%
  select(orthogroup, functCat) %>%
  distinct() %>%
  group_by(functCat) %>%
  summarise(n = n()) %>%
  left_join(T_functCat) %>%
  mutate(functCatDescription = str_c(functCat, " (n = ", n, ")", "\n", str_wrap(description, width= 55)))

# do pcoa per functional category and add genome info
T_pcoa2 <- T_pcoa2 %>%
  split(f = .$functCat) %>%
  lapply(FUN = function(table) {
    if (nrow(table) <= 5) return(NULL)
    functionalCategory = table$functCat[1]
    make_pcoaTable(table) %>%
      mutate(functCat = functionalCategory) %>%
      mutate(V1 = V1 - (min(V1) + max(V1)) / 2,
             V2 = V2 - (min(V2) + max(V2)) / 2) %>%     
      return()
  }) %>% bind_rows() %>%
  left_join(T_genome)

# add number of orthogroups per functional category info
T_pcoa2 <- left_join(T_pcoa2, T_nOrthogroupsPerFunctCat)

# print categories not included in pcoa plot
T_pcoa2 %>%
  select(functCat, n) %>%
  distinct() %>%
  arrange(n)

# remove functional categories with too few orthogroups
T_pcoa2 <- filter(T_pcoa2, n > 20)

# make and save plot! 
ggplot(T_pcoa2, aes(x = V1, y = V2, col = clade)) +
  geom_point(alpha=0.8) +
  xlim(c(-0.3, 0.3)) + ylim(c(-0.3, 0.3)) +
  scale_color_manual(values = cladeColors) +
  facet_wrap(~ functCatDescription, ncol = 4) +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
ggsave(filename = "results/pcoa_perFunctCat.png", height = 40, width = 35, units = "cm")
```

## Number of orthogroups per genome, by functional category

```{r}
# make table with genome, orthogroup, ngenes and funct cat
# remark: mutate step is to make sure that orthogroups without
# eggnogFam hit get functional category "S" (function unknown)
T_plot <- left_join(V_genome_orthogroup, T_orthogroup) %>%
  left_join(T_eggnogFam_functCat) %>%
  mutate(functCat = ifelse(is.na(functCat), "S", functCat)) %>%
  left_join(T_nOrthogroupsPerFunctCat) %>%
  select(genome, orthogroup, ngenes, functCatDescription) %>%
  left_join(T_genome) %>%
  group_by(functCatDescription, clade, species, genome) %>%
  summarise(nOrthogroups = n(), nGenes = sum(ngenes)) %>%
  ungroup()

ggplot(T_plot, aes(x = clade, y = nGenes, col = clade)) +
  geom_jitter(width = 0.2) +
  expand_limits(y = 0) +
  scale_color_manual(values = cladeColors) +
  facet_wrap(~ functCatDescription, ncol = 4, scales = "free_y") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'))
ggsave(filename = "results/nGenes_perFunctCat.png", height = 40, width = 35, units = "cm")
```

